\textbf{Linear Algebra -- Part II}\\A summary for MIT 18.06SC

### 2020-07-23 skills/math/linear_algebra/linear_algebra_summary_p2.tex
upload

@@

Let $\mA \in \bbR^{m \times n}$ be a matrix. There are four subspaces related to $\mA\colon$ column space $C(\mA)$, row space $C(\mA^T)$, nullspace $N(\mA)$, and left nullspace $N(\mA^T)$. Suppose the rank of $\mA$ is $rank(\mA) = r$. The dimensions of these subspaces are: 
\[ dim(C(\mA)) = dim(C(\mA^T)) = r,\quad  dim(N(A)) = n-r,\quad dim(N(\mA^T)) = m-r. \]

@@

\begin{thm}[Relationship between $\mA^T \mA$ and $\mA$]\label{thm:aa}Let $\mA \in \bbR^{m \times n}$ be a matrix. We have
	\begin{equation}
		N(\mA^T \mA) = N(\mA) \quad \text{and} \quad rank(\mA^T \mA) = rank(\mA).
	\end{equation}
\end{thm}
\begin{proof}
	First, we prove that $N(\mA^T \mA) = N(\mA)$.

	On one hand, if the vector $x \in N(\mA)$, then $\mA x = 0$. Multiplying the matrix $\mA^T$ on both sides of the equation, we have $ \mA^T  \mA x = 0$. Therefore, for any $x \in N(\mA)$, the vector $x \in N(\mA^T \mA)$. 
	\iffalse

	On the other hand, we  prove  by contradiction that for any $x \in N(\mA^T \mA)$ the vector $x \in N(\mA)$.

	Suppose there is a vector $x \in N(\mA^T \mA^T)$ but $x \notin N(\mA)$. We have
	\begin{equation}\label{eq:aaa}
		\mA x = b \neq 0, \quad \mA^T \mA x = 0 \quad \Rightarrow \quad \mA^T b = 0.
	\end{equation} 
	By the first equality in \eqref{eq:aaa}, $b \in C(\mA)$, and by the third equation in \eqref{eq:aaa}, $ b \in N(\mA^T)$. This contradicts the Theorem~\ref{thm:ortho} that $C(\mA) \perp N(\mA^T)$. Therefore, for any $x \in N(\mA^T \mA)$, the vector $x \in N(\mA)$.

@@

	On the other hand, if the vector $x \in N(\mA^T \mA^T)$, we have
	\[ \mA^T \mA x = 0.\]
	Then, the vector $\mA x \in N(\mA^T)$. By the definition of matrix-to-vector multiplication, we also have $\mA x \in C(\mA)$.  By the Theorem~\ref{thm:ortho}, the subspaces $N(\mA^T) \perp C(\mA)$. Therefore, $\mA x = 0$, and thus $x \in N(\mA)$.
	\vspace{0.2cm}

	Next, we prove that $rank(\mA^T \mA) = rank(\mA)$.

	Given $N(\mA^T \mA) = N(\mA)$, we have $rank(\mA^T \mA) = n - dim(N(\mA^T \mA)) = n - dim(N(\mA)) = rank(\mA)$.
\end{proof}

\begin{cor}[Invertibility of $\mA^T \mA$]\label{cor:invert}
	If $\mA$ has independent columns, then $\mA^T \mA$ is invertible.
\end{cor}
\begin{proof}
Suppose $\mA \in \bbR^{m \times n}$ is a matrix with independent columns; i.e., $rank(\mA) = n$. By Theorem~\ref{thm:aa}, $rank(\mA^T \mA) = rank(\mA) = n$. Since $\mA^T \mA \in \bbR^{n \times n}$ is a square matrix, $\mA^T \mA$ is invertible.
\end{proof}

\subsection{Projections onto subspaces}
\begin{defn}[\textit{Projection and projection matrix}]\label{def:project} Let  $x \in \bbR^{x}$ be a vector and $\mA^{n \times m}$ be a matrix. The vector $p_{C(\mA)}(x) \in C(\mA)$ that satisfies
	\begin{equation}\label{eq:project}
		(x-p_{C(\mA)}(x)) \perp C(\mA),
	\end{equation} 
	is called the projection of vector $x$ onto the column space $C(\mA)$. For all $x \in \bbR^n$, the matrix  $\mP_{C(\mA)} \in \bbR^{n \times n}$  that satisfies	\begin{equation}
		p_{C(\mA)}(x) =\mP_{C(\mA)} x,
	\end{equation}
	is called the projection matrix of $\mA$ from $\bbR^n$ onto the column space  $C(\mA)$. Note that the projection $p_{C(\mA)}(x)$ depends on the vector $x$ and the space $C(\mA)$, while the projection matrix $\mP_{C(\mA)}$ depends only on the space $C(\mA)$. 
\end{defn}

\begin{prop}[Projection matrix of $C(\mA)$]\label{prop:project}
	Let  $x \in \bbR^{n}$ be a vector and $\mA^{n \times m}$ be a matrix with independent columns. The projection matrix of $\mA$ from $\bbR^n$ onto the column space  $C(\mA)$ is 
	\begin{equation}
		\mP_{C(\mA)} =  \mA (\mA^T \mA)^{-1} \mA^T.
	\end{equation}
\end{prop}

\begin{proof}
Let $x_0$ be an arbitrary vector of $\bbR^n$. By Definition~\ref{def:project}, the projection $p_{C(\mA)}(x_0) \in C(\mA)$. Then, there exists a vector $v(x_0) \in \bbR^m$ such that $p_{C(\mA)}(x_0) = \mA  v(x_0)$.  By equation~\eqref{eq:project}, we have 
\begin{equation}\label{eq:projectca}
	A^T (x_0 - \mA v(x_0)) = 0 \quad \Rightarrow  \quad v(x_0) = (\mA^T \mA)^{-1} \mA^T x_0 \quad \Rightarrow \quad p_{C(\mA)}(x_0)  = \mA (\mA^T \mA)^{-1} \mA^T  x_0. 
\end{equation}
Since $x_0$ is arbitrary, the equation~\eqref{eq:projectca} still holds after replacing all the $x_0$ by $x$, for all $x \in \bbR^n$. Therefore, the matrix $\mP_{C(\mA)} =  \mA (\mA^T \mA)^{-1} \mA^T$ is the projection matrix of $\mA$ from $\bbR^n$ onto the column space $C(\mA)$.
\end{proof}

\begin{thm}[Properties of projection matrix]\label{thm:projectm}
Let $\mA \in \bbR^{m \times n}$ be a matrix, and $\mP_{C(\mA)}$ be the projection matrix of $\mA$ onto the column space $C(\mA)$. Then,
	\begin{equation}\label{eq:projectm}
		\mP_{C(\mA)}^T = \mP_{C(\mA)}\quad \text{and}\quad  \mP_{C(\mA)}^2 = \mP_{C(\mA)}.
	\end{equation}
\end{thm}
\begin{proof}
	By Proposition~\ref{prop:project}, the projection matrix is $\mP_{C(\mA)} =   \mA(\mA^T \mA)^{-1} \mA^T$. Then,
	\[  \mP_{C(\mA)}^T = (\mA(\mA^T \mA)^{-1} \mA^T )^T = \mA (\mA^T \mA)^{-1} \mA^T = \mP_{C(\mA)}. \]
Similarly, we have
	\[  \mP_{C(\mA)}^2 = \mP_{C(\mA)}^T \mP_{C(\mA)} =  \mA (\mA^T \mA)^{-1} \mA^T  \mA (\mA^T \mA)^{-1} \mA^T =  \mA (\mA^T \mA)^{-1} \mA^T = \mP_{C(\mA)}. \]
\end{proof}

\begin{cor}[Projection onto $N(\mA^T)$]\label{cor:iprojectm}
	Suppose $\mP_{C(\mA)}$ is the projection matrix of $\mA$ from $\bbR^n$ onto the column space $C(\mA)$. Then $\mI-\mP_{C(\mA)}$ is the projection matrix of $\mA$ from $\bbR^n$ onto the left nullspace $N(\mA^T)$. 
\end{cor}

\begin{proof}
	By Definition~\ref{def:project}, for all $x \in \bbR^n$, $\mP_{C(\mA)} x$ is the projection of $x$ from $\bbR^n$ to $C(\mA)$. Then,
	\[ x - \mP_{C(\mA)} x \perp C(\mA) \quad \Rightarrow \quad (\mI-\mP_{C(\mA)})x \perp C(\mA). \]
	By Theorem~\ref{thm:ortho}, we have $C(\mA) \perp N(\mA^T)$. Since $rank(C(\mA)) + rank(N(\mA^T)) = n$, $C(\mA)  \cup N(\mA^T) = \bbR^n$. Then, we have
	\[ (\mI-\mP_{C(\mA)})x \in N(\mA^T).\]
	Since $\mP_{C(\mA)} x$ is the projection of $x$ from $\bbR^n$ to $C(\mA)$,  the vector $\mP_{C(\mA)} x \in C(\mA)$, for all  $x \in \bbR^n$. We have,
	\[ x -  (\mI-\mP_{C(\mA)})x = \mP_{C(\mA)} x \in C(\mA) \quad \Rightarrow \quad  x -  (\mI-\mP_{C(\mA)})x \perp N(\mA^T).\]
	Therefore, for all $x \in \bbR^n$, $(\mI-\mP_{C(\mA)})x$ is the projection of $x$ from $\bbR^n$ onto the left nullspace $N(\mA^T)$, and thus $\mI-\mP_{C(\mA)}$ is the projection matrix of $\mA$ from $\bbR^n$ onto $N(\mA^T)$.
\end{proof}

\subsection{Projection matrices and least squares}
Let $y \in \bbR^n$ be a vector and $\mX \in \bbR^{n \times (k+1)}$ be a design matrix. We propose the linear regression model as
\[ y = \mX \beta +\epsilon,\]
where $\beta = (\beta_0,\beta_1,...,\beta_k)^T$ are regression coefficients and $\epsilon$ is the noise. The least-squares estimate of $\beta$ is the minimizer of the squared loss; i.e.,
\[  \hat \beta_{LS} = \argmin_{\beta \in \bbR^{k+1}} \onorm{y - \mX \beta}^2, \]
where $\onorm{\cdot}$ is the Euclidean norm.  The projection of $y$ onto $C(\mX)$ minimizes the Euclidean distance from $y$ to the column space $C(\mX)$. Hence,   we consider the vector $\mX 
\hat \beta_{LS}$ as the projection of $y$ onto the column space $C(\mX)$. Therefore, we may use projection to solve the minimization problem.

@@

Suppose $\mQ \in \bbR^{m \times n}$ is an orthonormal matrix. Then, we have $\mQ^T \mQ = \mI_n$.  When $m = n$, $\mQ$ is an orthogonal matrix and $\mQ^T = \mQ^{-1}$. The projection matrix of $\mQ$ onto the column space $C(\mQ)$, denoted $\mP_{C(\mQ)}$,  becomes $\mP_{C(\mQ)} = \mI_m$.

\begin{defn}[\textit{Gram-Schimidt Process and QR decomposition}]\label{gram-schimidt}
Let $\mA \in \bbR^{m \times n}$ be a matrix with $rank(\mA) = n$, and  $a_i \in \bbR^{m}$ be the $i$-th column of matrix $\mA$, for all $i \in [n]$. Gram-Schimidt process finds the orthonormal basis for $C(\mA)$ as follows,
\begin{align}
	&u_1 = a_1, &&\quad e_1 = \frac{u_1}{\onorm{u_1}};\\
	&u_2 = a_2 - \frac{u_1^T a_2}{u_1^T u_1} u_1 ,&&\quad  e_2 = \frac{u_2}{\onorm{u_2}};\\
	&u_3 = a_3 - \frac{u_1^T a_3}{u_1^T u_1} u_1 - \frac{u_2^T a_3}{u_2^T u_2} u_2,&&\quad  e_2 = \frac{u_3}{\onorm{u_3}};\\
	\vdots\\
	&u_n = a_n - \sum_{k = 1}^{n-1} \frac{u_k^T a_k}{u_k^T u_k} u_k, && \quad e_n = \frac{u_n}{\onorm{u_n}}.
\end{align}
 The vectors $e_1,...,e_n$ are orthonormal basis of $C(\mA)$. Based on the basis obtained by Gram-Schimidit, we decompose the matrix $\mA$ as follows,
 \begin{equation}\label{eq:qr}
 	\mA = [a_1,...,a_n] = [e_1,....,e_2] \begin{bmatrix}
 		e_1^T a_1 & e_1^T a_2 & \cdots & e_1^T a_n\\
 		0&e_2^T a_2 & \cdots & e_2^T a_n\\
 		\vdots &\vdots & & \vdots \\
 		0 &0&\cdots & e_n^T a_n
 	\end{bmatrix} = \mQ \mR,
 \end{equation}
	where $\mQ \in \bbR^{m \times n}$ is an orthonormal matrix and $\mR \in \bbR^{n \times n}$ is an upper triangular matrix. The matrix decomposition in equation~\eqref{eq:qr} is called QR decomposition.
\end{defn}

@@

\begin{defn}[\textit{Parallelepiped}]
	The parallelepiped determined by $n$ vectors $v_1,...,v_n \in \bbR^n$ is defined as the following subset,
	\[ P = \{ a_1 v_1 + \cdots + a_n v_n \colon 0 \leq a_1,...,a_n \leq 1 \}. \]
	 We use $vol(P)$ to denote the volume of the parallelepiped  $P$. 
\end{defn}

\begin{prop}[Determinants and volumes]
The absolute determinant of a square matrix $\mA$ is the volume of parallelepiped determined by the rows of $\mA$; i.e.,
	\[ |\det(\mA)| = vol(P(\mA)), \]
	where $P(\mA)$ is the parallelepiped determined by the rows of $\mA$.
\end{prop}

@@

\begin{defn}[\textit{Trace of square matrix}]\label{def:trace}
Let $\mA = \entry{a_{ij}} \in \bbR^{n \times n}$ be a square matrix. The trace of $\mA$ is defined as
\[ tr(\mA) = \sum_{i = 1}^n a_{ii}.\] 
\end{defn}

\begin{defn}[\textit{Characteristic polynomial}]\label{def:charac}
	Let $\lambda$ denote an eigenvalue of the matrix $\mA\in \bbR^{n \times n}$.
	The determinant of $\mA - \lambda \mI_n$ is a polynomial of $\lambda$, denoted $P(\lambda)$. We call the polynomial $P(\lambda)$ as the characteristic polynomial of $\mA$. Specifically,
	\begin{equation}\label{def:charapoly}
		 P(\lambda)  = (-1)^n \lambda^n +  (-1)^n  tr(\mA)\lambda^{n-1} + \cdots + (-1)^n \det(\mA). 
	\end{equation}
\end{defn}

@@

\begin{thm}[Summation and production of eigenvalues]\label{thm:speigen}
	Let  $\mA \in \bbR^{n \times n}$ be a square matrix, and $\lambda_1,...\lambda_n$ be the eigenvalues of $\mA$. Then,
	\begin{equation}
		\sum_{i=1}^n \lambda_i = tr(\mA)\quad \text{and} \quad \prod_{i=1}^n \lambda_i = \det(\mA).
	\end{equation} 
\end{thm}

\begin{proof}
	First, we re-write the characteristic polynomial $P(\lambda)$ of $\mA$ as a factor formula,
	\begin{equation}\label{eq:poly}
		P(\lambda) = (-1)^n (\lambda - \lambda_1)\cdots (\lambda - \lambda_n), 
	\end{equation}   
	where $\lambda_i$ are the eigenvalues of $\mA$, for all $ i \in [n]$. In equation~\eqref{eq:poly}, the coefficient for $\lambda^{n-1}$ is equal to $(-1)^n\sum_{i=1}^n \lambda_i$. Comparing the equation~\eqref{eq:poly} with Definition~\ref{def:charac}, we obtain $tr(\mA) = \sum_{i=1}^n \lambda_i$. Similarly, the constant term in equation~\eqref{eq:poly} is $(-1)^n \prod_{i=1}^n \lambda_i $, while the constant term in Defintion~\ref{def:charac} is $(-1)^n \det(\mA)$.  Therefore, we obtain $\prod_{i=1}^n \lambda_i = \det(\mA)$.
\end{proof}

>>>>

\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\vol}{vol}
\DeclareMathOperator{\tr}{tr}

@@

Let $\mA \in \bbR^{m \times n}$ be a matrix. There are four subspaces related to $\mA\colon$ column space $C(\mA)$, row space $C(\mA^T)$, nullspace $N(\mA)$, and left nullspace $N(\mA^T)$. Suppose the rank of $\mA$ is $\rank(\mA) = r$. The dimensions of these subspaces are: 
\[ \dim(C(\mA)) = \dim(C(\mA^T)) = r,\quad  \dim(N(A)) = n-r,\quad \dim(N(\mA^T)) = m-r. \]

@@

For notational convenience, we use the shorthands $\mA$ for the column space $C(\mA)$ and $\mA^{\perp}$ for the nullspace $N(\mA)$ in the rest of this section. Though $\mA$ usually denotes a matrix, the meaning of $\mA$ should be clear given the context. For example, $\rank(\mA)$ refers the rank of matrix $\mA$, while $\dim(\mA)$ refers to the dimension of the column space $C(\mA)$. Specifically, for a vector $v$, $v \in \mA$ means that the vector $v$ belongs to the column space $C(\mA)$.

\vspace{0.2cm}
Given a matrix $\mA \in \bbR^{m \times n}$, we also rewrite the Theorem~\ref{thm:ortho} as
\[ \mA^T \perp \mA^{\perp} \quad \text{and} \quad \mA \perp \mA^{T,\perp}. \]

\begin{thm}[Relationship between $\mA^T \mA$ and $\mA$]\label{thm:aa}Let $\mA \in \bbR^{m \times n}$ be a matrix. We have
	\begin{equation}
		(\mA^T \mA)^{\perp} = \mA^{\perp} \quad \text{and} \quad \rank(\mA^T \mA) = \rank(\mA).
	\end{equation}
\end{thm}
\begin{proof}
	First, we prove that $(\mA^T \mA)^{\perp} = \mA^{\perp}$.

	On one hand, if the vector $x \in \mA^{\perp}$, then $\mA x = 0$. Multiplying the matrix $\mA^T$ on both sides of the equation, we have $ \mA^T  \mA x = 0$. Therefore, for any $x \in \mA^{\perp}$, the vector $x \in (\mA^T \mA)^{\perp}$. 
	\iffalse

	On the other hand, we  prove  by contradiction that for any $x \in (\mA^T \mA)^{\perp}$ the vector $x \in \mA^{\perp}$.

	Suppose there is a vector $x \in (\mA^T \mA^T)^{\perp}$ but $x \notin \mA^{\perp}$. We have
	\begin{equation}\label{eq:aaa}
		\mA x = b \neq 0, \quad \mA^T \mA x = 0 \quad \Rightarrow \quad \mA^T b = 0.
	\end{equation} 
	By the first equality in \eqref{eq:aaa}, $b \in \mA$, and by the third equation in \eqref{eq:aaa}, $ b \in \mA^{T,\perp}$. This contradicts the Theorem~\ref{thm:ortho} that $\mA \perp \mA^{T,\perp}$. Therefore, for any $x \in (\mA^T \mA)^{\perp}$, the vector $x \in \mA^{\perp}$.

@@

	On the other hand, if the vector $x \in (\mA^T \mA)^{\perp}$, we have
	\[ \mA^T \mA x = 0.\]
	Then, the vector $\mA x \in  \mA^{T,\perp}$. By the definition of matrix-to-vector multiplication, we also have $\mA x \in \mA$.  By the Theorem~\ref{thm:ortho}, the subspaces $\mA^{T,\perp} \perp \mA$. Therefore, $\mA x = 0$, and we obtain that $x \in \mA^{\perp}$.
	\vspace{0.2cm}

	Next, we prove that $\rank(\mA^T \mA) = \rank(\mA)$.

	Given $(\mA^T \mA)^{\perp} = \mA^{\perp}$, we have $\rank(\mA^T \mA) = n - \dim((\mA^T \mA)^{\perp}) = n - \dim(\mA^{\perp}) = \rank(\mA)$.
\end{proof}

\begin{cor}[Invertibility of $\mA^T \mA$]\label{cor:invert}
	If $\mA$ has independent columns, then $\mA^T \mA$ is invertible.
\end{cor}
\begin{proof}
Suppose $\mA \in \bbR^{m \times n}$ is a matrix with independent columns; i.e., $\rank(\mA) = n$. By Theorem~\ref{thm:aa}, $\rank(\mA^T \mA) = \rank(\mA) = n$. Since $\mA^T \mA \in \bbR^{n \times n}$ is a square matrix, $\mA^T \mA$ is invertible.
\end{proof}

\subsection{Projections onto subspaces}
\begin{defn}[\textit{Projection and projection matrix}]\label{def:project} Let  $x \in \bbR^{x}$ be a vector and $\mA^{n \times m}$ be a matrix. The vector $p_{\mA}(x) \in \mA$ that satisfies
	\begin{equation}\label{eq:project}
		(x-p_{\mA}(x)) \perp \mA,
	\end{equation} 
	is called the projection of vector $x$ onto the column space $\mA$. For all $x \in \bbR^n$, the matrix  $\mP_{\mA} \in \bbR^{n \times n}$  that satisfies	\begin{equation}
		p_{\mA}(x) =\mP_{\mA} x,
	\end{equation}
	is called the projection matrix of $\mA$ from $\bbR^n$ onto the column space $\mA$. Note that the projection $p_{\mA}(x)$ depends on the vector $x$ and the space $\mA$, while the projection matrix $\mP_{\mA}$ depends on the space $\mA$ only. 
\end{defn}

\begin{prop}[Projection matrix of $\mA$]\label{prop:project}
	Let  $x \in \bbR^{n}$ be a vector and $\mA^{n \times m}$ be a matrix with independent columns. The projection matrix of $\mA$ from $\bbR^n$ onto the column space  $\mA$ is 
	\begin{equation}
		\mP_{\mA} =  \mA (\mA^T \mA)^{-1} \mA^T.
	\end{equation}
\end{prop}

\begin{proof}
Let $x_0$ be an arbitrary vector of $\bbR^n$. By Definition~\ref{def:project}, the projection $p_{\mA}(x_0) \in \mA$. Then, there exists a vector $v(x_0) \in \bbR^m$ such that $p_{\mA}(x_0) = \mA  v(x_0)$.  By equation~\eqref{eq:project}, we have 
\begin{equation}\label{eq:projectca}
	A^T (x_0 - \mA v(x_0)) = 0 \quad \Rightarrow  \quad v(x_0) = (\mA^T \mA)^{-1} \mA^T x_0 \quad \Rightarrow \quad p_{\mA}(x_0)  = \mA (\mA^T \mA)^{-1} \mA^T  x_0. 
\end{equation}
Since $x_0$ is arbitrary, the equation~\eqref{eq:projectca} still holds after replacing all the $x_0$ by $x$, for all $x \in \bbR^n$. Therefore, the matrix $\mP_{\mA} =  \mA (\mA^T \mA)^{-1} \mA^T$ is the projection matrix of $\mA$ from $\bbR^n$ onto the column space $\mA$.
\end{proof}

\begin{thm}[Properties of projection matrix]\label{thm:projectm}
Let $\mA \in \bbR^{m \times n}$ be a matrix, and $\mP_{\mA}$ be the projection matrix of $\mA$ onto the column space $\mA$. Then,
	\begin{equation}\label{eq:projectm}
		\mP_{\mA}^T = \mP_{\mA}\quad \text{and}\quad  \mP_{\mA}^2 = \mP_{\mA}.
	\end{equation}
\end{thm}
\begin{proof}
	By Proposition~\ref{prop:project}, the projection matrix is $\mP_{\mA} =   \mA(\mA^T \mA)^{-1} \mA^T$. Then,
	\[  \mP_{\mA}^T = (\mA(\mA^T \mA)^{-1} \mA^T )^T = \mA (\mA^T \mA)^{-1} \mA^T = \mP_{\mA}. \]
Similarly, we have
	\[  \mP_{\mA}^2 = \mP_{\mA}^T \mP_{\mA} =  \mA (\mA^T \mA)^{-1} \mA^T  \mA (\mA^T \mA)^{-1} \mA^T =  \mA (\mA^T \mA)^{-1} \mA^T = \mP_{\mA}. \]
\end{proof}

\begin{cor}[Projection onto $\mA^{T,\perp}$]\label{cor:iprojectm}
	Suppose $\mP_{\mA}$ is the projection matrix of $\mA$ from $\bbR^n$ onto the column space $\mA$. Then $\mI-\mP_{\mA}$ is the projection matrix of $\mA$ from $\bbR^n$ onto the left nullspace $\mA^{T,\perp}$. 
\end{cor}

\begin{proof}
	By Definition~\ref{def:project}, for all $x \in \bbR^n$, $\mP_{\mA} x$ is the projection of $x$ from $\bbR^n$ onto $\mA$. Then,
	\[ x - \mP_{\mA} x \perp \mA \quad \Rightarrow \quad (\mI-\mP_{\mA})x \perp \mA. \]
	By Theorem~\ref{thm:ortho}, we have $\mA \perp \mA^{T,\perp}$. Since $\dim(\mA) + \dim(\mA^{T,\perp}) = n$, we have $\mA  \cup \mA^{T,\perp} = \bbR^n$. Then, we have
	\[ (\mI-\mP_{\mA})x \in  \mA^{T,\perp}.\]
	Since $\mP_{\mA} x$ is the projection of $x$ from $\bbR^n$ to $\mA$,  the vector $\mP_{\mA} x \in \mA$, for all  $x \in \bbR^n$. We have,
	\[ x -  (\mI-\mP_{\mA})x = \mP_{\mA} x \in \mA \quad \Rightarrow \quad  x -  (\mI-\mP_{\mA})x \perp \mA^{T,\perp}.\]
	Therefore, for all $x \in \bbR^n$, $(\mI-\mP_{\mA})x$ is the projection of $x$ from $\bbR^n$ onto the left nullspace $\mA^{T,\perp}$, and $\mI-\mP_{\mA}$ is the projection matrix of $\mA$ from $\bbR^n$ onto $\mA^{T,\perp}$.
\end{proof}

\subsection{Projection matrices and least squares}
Let $y \in \bbR^n$ be a vector and $\mX \in \bbR^{n \times (k+1)}$ be a design matrix. We propose the linear regression model as
\[ y = \mX \beta +\epsilon,\]
where $\beta = (\beta_0,\beta_1,...,\beta_k)^T$ are regression coefficients and $\epsilon$ is the noise. The least-squares estimate of $\beta$ is the minimizer of the squared loss; i.e.,
\[  \hat \beta_{LS} = \argmin_{\beta \in \bbR^{k+1}} \onorm{y - \mX \beta}^2, \]
where $\onorm{\cdot}$ is the Euclidean norm.  The projection of $y$ onto $\mX$ minimizes the Euclidean distance from $y$ to the column space $\mX$. Hence,   we consider the vector $\mX 
\hat \beta_{LS}$ as the projection of $y$ onto the column space $\mX$. Therefore, we may use projection to solve the minimization problem.

@@

Suppose $\mQ \in \bbR^{m \times n}$ is an orthonormal matrix. Then, we have $\mQ^T \mQ = \mI_n$.  When $m = n$, $\mQ$ is an orthogonal matrix and $\mQ^T = \mQ^{-1}$. The projection matrix of $\mQ$ onto the column space $C(\mQ)$, denoted $\mP_{\mQ}$,  becomes $\mP_{\mQ} = \mI_m$.

\begin{defn}[\textit{Gram-Schimidt Process and QR decomposition}]\label{gram-schimidt}
Let $\mA \in \bbR^{m \times n}$ be a matrix with $\rank(\mA) = n$, and  $a_i \in \bbR^{m}$ be the $i$-th column of matrix $\mA$, for all $i \in [n]$. Gram-Schimidt process finds the orthonormal basis for $\mA$ as follows,
\begin{align}
	&u_1 = a_1, &&\quad e_1 = \frac{u_1}{\onorm{u_1}};\\
	&u_2 = a_2 - \frac{u_1^T a_2}{u_1^T u_1} u_1 ,&&\quad  e_2 = \frac{u_2}{\onorm{u_2}};\\
	&u_3 = a_3 - \frac{u_1^T a_3}{u_1^T u_1} u_1 - \frac{u_2^T a_3}{u_2^T u_2} u_2,&&\quad  e_2 = \frac{u_3}{\onorm{u_3}};\\
	\vdots\\
	&u_n = a_n - \sum_{k = 1}^{n-1} \frac{u_k^T a_k}{u_k^T u_k} u_k, && \quad e_n = \frac{u_n}{\onorm{u_n}}.
\end{align}
 The vectors $e_1,...,e_n$ are orthonormal basis of $\mA$. Based on the basis obtained by Gram-Schimidit, we decompose the matrix $\mA$ as follows,
 \begin{equation}\label{eq:qr}
 	\mA = [a_1,...,a_n] = [e_1,....,e_2] \begin{bmatrix}
 		e_1^T a_1 & e_1^T a_2 & \cdots & e_1^T a_n\\
 		0&e_2^T a_2 & \cdots & e_2^T a_n\\
 		\vdots &\vdots & & \vdots \\
 		0 &0&\cdots & e_n^T a_n
 	\end{bmatrix} = \mQ \mR,
 \end{equation}
	where $\mQ \in \bbR^{m \times n}$ is an orthonormal matrix and $\mR \in \bbR^{n \times n}$ is an upper triangular matrix. The matrix decomposition in equation~\eqref{eq:qr} is called QR decomposition.
\end{defn}

@@

\begin{defn}[\textit{Parallelepiped}]
	The parallelepiped determined by $n$ vectors $v_1,...,v_n \in \bbR^n$ is defined as the following subset,
	\[ P = \{ a_1 v_1 + \cdots + a_n v_n \colon 0 \leq a_1,...,a_n \leq 1 \}. \]
	 We use $\vol(P)$ to denote the volume of the parallelepiped  $P$. 
\end{defn}

\begin{prop}[Determinants and volumes]
The absolute determinant of a square matrix $\mA$ is the volume of parallelepiped determined by the rows of $\mA$; i.e.,
	\[ |\det(\mA)| = \vol(P(\mA)), \]
	where $P(\mA)$ is the parallelepiped determined by the rows of $\mA$.
\end{prop}

@@

\begin{defn}[\textit{Trace of square matrix}]\label{def:trace}
Let $\mA = \entry{a_{ij}} \in \bbR^{n \times n}$ be a square matrix. The trace of $\mA$ is defined as
\[ \tr(\mA) = \sum_{i = 1}^n a_{ii}.\] 
\end{defn}

\begin{defn}[\textit{Characteristic polynomial}]\label{def:charac}
	Let $\lambda$ denote an eigenvalue of the matrix $\mA\in \bbR^{n \times n}$.
	The determinant of $\mA - \lambda \mI_n$ is a polynomial of $\lambda$, denoted $P(\lambda)$. We call the polynomial $P(\lambda)$ as the characteristic polynomial of $\mA$. Specifically,
	\begin{equation}\label{def:charapoly}
		 P(\lambda)  = (-1)^n \lambda^n +  (-1)^n  \tr(\mA)\lambda^{n-1} + \cdots + (-1)^n \det(\mA). 
	\end{equation}
\end{defn}

@@

\begin{thm}[Summation and production of eigenvalues]\label{thm:speigen}
	Let  $\mA \in \bbR^{n \times n}$ be a square matrix, and $\lambda_1,...\lambda_n$ be the eigenvalues of $\mA$. Then,
	\begin{equation}
		\sum_{i=1}^n \lambda_i = \tr(\mA)\quad \text{and} \quad \prod_{i=1}^n \lambda_i = \det(\mA).
	\end{equation} 
\end{thm}

\begin{proof}
	First, we re-write the characteristic polynomial $P(\lambda)$ of $\mA$ as a factor formula,
	\begin{equation}\label{eq:poly}
		P(\lambda) = (-1)^n (\lambda - \lambda_1)\cdots (\lambda - \lambda_n), 
	\end{equation}   
	where $\lambda_i$ are the eigenvalues of $\mA$, for all $ i \in [n]$. In equation~\eqref{eq:poly}, the coefficient for $\lambda^{n-1}$ is equal to $(-1)^n\sum_{i=1}^n \lambda_i$. Comparing the equation~\eqref{eq:poly} with Definition~\ref{def:charac}, we obtain $\tr(\mA) = \sum_{i=1}^n \lambda_i$. Similarly, the constant term in equation~\eqref{eq:poly} is $(-1)^n \prod_{i=1}^n \lambda_i $, while the constant term in Defintion~\ref{def:charac} is $(-1)^n \det(\mA)$.  Therefore, we obtain $\prod_{i=1}^n \lambda_i = \det(\mA)$.
\end{proof}
